Data Exploration

### 2019-01-17 src/biological_data_prep.Rmd
datasets merged, vif run, finished with missing data stuff

```{r}
names(data_aea)[names(data_aea)=="freq"] <- "total_cell_obs"
head(data_aea)
```

time-sliced
```{r}
data_aea <- merge(data_aea, obs_cell_yymm, by.x=c("cell_id", "year", "month"), by.y=c("cell_id", "year", "month"))
names(data_aea)[names(data_aea)=="freq"] <- "yymm_cell_obs"
head(data_aea)
```

@@

chl_depth = 2394 missing values
o2_depth = 2394 missing values
salinity_depth = 3133 missing values
temp_depth = 3109 missing values

This is interesting...
chl and o2 are missing the same number of points - they come from the biomer dataset; salinity and temp come from glorys and have different values.

@@

```{r}
presence_wdepth2 <-  presence_wdepth[!complete.cases(presence_wdepth[28:37]), ]
nrow(presence_wdepth2) #3349
```
ok so there are 3349 points without any data.let's map these

@@

ok so 2014_07 and 1999_07 are both missing all data. Re-run the loop for these two time-periods (just add the nectcdfs for these files in renetcdf folder).

@@

lets see if that made any difference...

@@

so there are 2178 out of 3349 observations that are missing all data for depth. so what about the remaining 1171 points?

want to remove from the observations in presence_wdepthonly (all 3349 observations) the ones that appear in presence_wdepthonly_allmissing (2178 observations)

some code from [stack exchange](https://stackoverflow.com/questions/10907359/r-remove-rows-from-one-data-frame-that-are-in-another)

```{r}
presence_wdepthonly_somemissing <-  rbind(presence_wdepthonly_allmissing, presence_wdepthonly)
presence_wdepthonly_somemissing <- presence_wdepthonly_somemissing[!duplicated(presence_wdepthonly_somemissing, fromLast = FALSE) & !duplicated(presence_wdepthonly_somemissing, fromLast = TRUE), ]
nrow(presence_wdepthonly_somemissing)
head(presence_wdepthonly_somemissing)
```

ok before going further, just rerun the loop again on all the data...

@@

It used to be
[1] 2394
[1] 2394
[1] 3133
[1] 3109
so actually a minor change (maybe this is because you added the dec 2015 data in - it wasn't available before). ok need to redo checks...

>>>>

```{r}
names(data_aea)[names(data_aea)=="freq"] <- "total_cell_obs_xy"
head(data_aea)
```

time-sliced
```{r}
data_aea <- merge(data_aea, obs_cell_yymm, by.x=c("cell_id", "year", "month"), by.y=c("cell_id", "year", "month"))
names(data_aea)[names(data_aea)=="freq"] <- "total_cell_obs_xyt"
head(data_aea)
```

@@

chl_depth = 2394 missing values
o2_depth = 2394 missing values
salinity_depth = 3109 missing values
temp_depth = 3109 missing values

This is interesting...
chl and o2 are missing the same number of points - they come from the biomer dataset; salinity and temp come from glorys and have the same number of points.

@@

```{r}
presence_wdepth2 <-  presence_wdepth[!complete.cases(presence_wdepth[28:37]), ]
nrow(presence_wdepth2) #3325
```
ok so there are 3325 points without any data.let's map these

@@

ok so 2014_07 and 1999_07 are both missing all data. Re-run the loop for this time-periods(just add the nectcdfs for these files in renetcdf folder).

@@

lets see if that made any difference... nope ok so it wasnt that i was missing the files

@@

ok before going further, just rerun the loop again on all the data just to make sure there wasn't some silly fail...

@@

It used to be
[1] 2394
[1] 2394
[1] 3109
[1] 3109
so no change.

ok well i think the answer here is that if i am going to use depth data, i should get rid of those with missing data. if i am going to use surface only data, then i can keep all points.

# glorys bathymetric data

so i just found that glorys also has a bathymetric data layer...

First you need to put the observations (data_aea) into a spatialpointsdataframe.
```{r}
xy <- data_aea[ ,c(19,20)] # This is to tell R where the coordinates are (in column 18 and 19). Note that the column order needs to be longitude, latitude
data_aea_sp <- SpatialPointsDataFrame(coords = xy, data = data_aea, proj4string = CRS("+proj=aea +lat_1=50 +lat_2=70 +lat_0=40 +lon_0=-60 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")) # The CRS is used here is for the albers equal area projection.
```

then covert the netcdf with the bathy layer (GLO-MFC_001_025_bathy.nclonlat.nc) to a brick and project to the albers equal area
```{r}
bathyrast <- raster("../data/env/GLO-MFC_001_025_bathy.nclonlat.nc") #because it is a single level file
aea <- raster("../output/env/aea.tif") 
bathyrast <- projectRaster(bathyrast, aea)
plot(bathyrast)
```

now extract and add to data_aea
```{r}
data_aea_sp$bottom_depth_glorys <- extract(x = bathyrast, y = data_aea_sp)
data_aea <- as.data.frame(data_aea_sp) # this coverts the 'spatialpoints dataframe' to a 'normal dataframe'
write.csv(data_aea, file = "../output/bio/presence_points_without_envdata_relooped_glbathy.csv", row.names = FALSE) # change the path to where ever you would like to save it
head(data_aea)
```

### 2019-01-17 src/data_exploration.Rmd
datasets merged, vif run, finished with missing data stuff

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
The "include=FALSE" supresses the package installation text appearing in the document...
```{r pre-install packages, include=FALSE}
packages <- c("plyr", "graphics", "ggplot2", "corrplot", "FSA") 
source("../src/ipak.R")
ipak(packages)
```

@@

```{r}
presence <- read.csv("../output/bio/presence_points_without_envdata_relooped.csv", header = TRUE)
head(presence)
```

@@

```{r}
background <- read.csv("../output/bio/background_complete_obs_cels.csv", header = TRUE)
head(background)
```

@@

>>>>

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
The "include=FALSE" supresses the package installation text appearing in the document...
```{r pre-install packages, include=FALSE}
packages <- c("plyr", "graphics", "ggplot2", "corrplot", "FSA", "car") 
source("../src/ipak.R")
ipak(packages)
```

@@

```{r}
presence <- read.csv("../output/bio/presence_points_without_envdata_relooped_glbathy.csv", header = TRUE)
head(presence)
```

@@

```{r}
background <- read.csv("../output/bio/background_complete_obs_cels_globot.csv", header = TRUE)
head(background)
```

@@

#vif

for this you need the joined dataset. this was created under mergig_datasets.src and the file is ../output/bio/presab.csv

```{r}
presab <- read.csv("../output/bio/presab.csv", header = TRUE)
```

```{r}
vif_allpresab <- vif(lm(occurrence ~ temp_surface + temp_depth + salinity_surface + salinity_depth + o2_surface + o2_depth + chl_surface + chl_depth + bottom_depth + mlp_surface + ssh_surface + gear + nao_sample + nao_prev + nao_winter + amo_sample + amo_prev + amo_winter, data = presab))
write.csv(vif_allpresab, "../output/bio/vif_allpresab.csv", row.names = TRUE)
vif_allpresab
```
interpret - see https://stats.stackexchange.com/questions/70679/which-variance-inflation-factor-should-i-be-using-textgvif-or-textgvif/96584#96584
To make GVIFs comparable across dimensions, we suggested using GVIF^(1/(2*Df)), where Df is the number of coefficients in the subset (ref fox and motette 1992 in zotero)
or the 2 continuous variables, GVIFˆ(1/(2*Df)) (which is basically the square root of the VIF/GVIF value as DF=1) is the proportional change of the standard error and confidence interval of their coefficients due to the level of collinearity. The GVIFˆ(1/(2*Df)) value of the categorical variable is a similar measure for the reduction in precision of the coefficients' estimation due to collinearity. 
apparently i just need to square GVIF^(1/(2*Df)) and then use the normal VIF "rule of thumb"...

```{r}
vif_allpresab_sq <- read.csv("../output/bio/vif_allpresab.csv", header = TRUE)
vif_allpresab_sq$GVIF2Dfsq <- vif_allpresab_sq$GVIF..1..2.Df..^2
write.csv(vif_allpresab_sq, "../output/bio/vif_allpresab_sq.csv", row.names = TRUE)
vif_allpresab_sq
```
